Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-22 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
working on the appendix

A general procedure to describe deviations to the QNM frequencies $\omega^{\rm GR}_{\ell m n}$ and
damping times $\tau^{\rm GR}_{\ell m n}$ of BHs of GR is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{\ell m n} &= \omega_{\ell m n}^{\rm GR} \, (1 + \delta \omega_{\ell m n}), \\
\tau_{\ell m n }   &= \tau_{\ell m  n}^{\rm GR}   \, (1 + \delta \tau_{\ell m n }),
\end{align}
\end{subequations}
%
where $\delta\omega_{\ell m n}$ and $\delta\tau_{\ell m  n}$ are the fractional deformation parameters,
$\ell$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization~\footnote{See also
Refs.~\cite{Glampedakis:2017dvb,Glampedakis:2019dqh,Silva:2019scu} and
Refs.~\cite{Cardoso:2019mqo,McManus:2019ulj,Kimura:2020mrh} for alternative parametrizations.}
was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Brito:2018rfr,Carullo:2018sfu,Ghosh:2021mrv}.

@@

The ParSpec framework was introduced in Ref.~\cite{Maselli:2019mjd} and can be
used to our purpose. It is an observable-based bivariate expansion of
Eq.~\eqref{eq:general_deviation}, given by
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{\ell m n} &= \frac{1}{M_{\rm f}} \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \omega^{(j)}_{\ell m n} \, \left( 1 + \gamma\, \delta \omega^{(j)}_{\ell m n} \right), \\
\tau_{\ell m n}   &= M_{\rm f}     \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \tau^{(j)}_{\ell m n}   \, \left( 1 + \gamma\, \delta \tau^{(j)}_{\ell m n} \right),
\end{align}
\end{subequations}
%
where $M_{\rm f}$ and $\chi_{\rm f}$ are the detector-frame final mass and spin, respectively;
the quantities $\omega_{\ell m n}^{(j)}$ and $\tau_{\ell m n}^{(j)}$ are dimensionless coefficients of the
expansion in spin for the QNMs of BHs in GR, while $\delta\omega_{\ell m n}^{(j)}$ and $\delta\tau_{\ell m n}^{(j)}$ are
source-independent dimensionless coefficients that characterize the corrections to the GR QNM at
each spin-order,
%
and $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence due to a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, which reads
%
\begin{equation}
\gamma
%
= \left(\frac{\ell_{\rm th}}{M^{\rm s}_{\rm f}}\right)^{p}
%
= \left[
\frac{\ell_{\rm th} c^2 (1 + z)}{G M_{\rm f}}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on the length-scale parameter $\ell_{\rm th}$ of the specific gravity theory
(non-GR modifications become important at distances $\lesssim \ell_{\rm th}$), and
the exponent $p$ is related to how the non-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\gamma$ dimensionless by \redsout{using} the \redsout{other}
length-scale associated with the remnant BH (i.e., its source-frame mass
$M^{\rm s}_{\rm f}$\footnote{Throughout this paper, we maintain the convention of
using the upperscript ``s" to denote source-frame quantities and plain baseline
symbols for detector-frame measurements.}), which we can also write in terms of the detector-frame mass $M_{\rm f}$
through the redshift $z$~\cite{Krolak:1987ofj}. Also, dividing by the factor $G/c^2$ allows us to
express $\ell_{\rm th}$ in physically intuitive metric units.

@@

We now describe the waveform model used in our paper to infer properties of a
BBH ringdown.
%
As in Refs.~\cite{Brito:2018rfr,Ghosh:2021mrv}, we use an IMR BBH waveform
model where the (complex-valued) frequencies describing the remnant object
are left additionally free and estimated directly from the data.

@@

Let us now establish the connection between the theory-independent framework of the
$\pSEOB$ waveform model and the theory-specific QNM calculations. In this paper we restrict ourselves
to the leading and next-to-leading order terms in the ParSpec expansion, as well as to the fundamental QNM
$(\ell, m, n) = (2,\, 2,\, 0)$. For this reason, for simplicity, we omit the subscripts hereafter
and rewrite $\omega_{\ell m n}$ and $\tau_{\ell m n}$, given by Eqs.~\eqref{eq:kerr_expansion} as,
%
\begin{subequations}
\label{eq:qnm_nongr_iso}
\begin{align}
    M_{\rm f}\, \omega &= \gamma \left [ \delta\omega^{(0)} \omega^{(0)} + \chi_{\rm f} \delta\omega^{(1)} \omega^{(1)} \right ]
    + \sum_{j=0}^{N_{\rm max}} \chi^{j}_{\rm f}\, \omega^{(j)} \,,
\\
    \frac{\tau}{M_{\rm f}}   &= \gamma \left [ \delta\tau^{(0)} \tau^{(0)} + \chi_{\rm f} \delta\tau^{(1)} \tau^{(1)} \right ]
    + \sum_{j=0}^{N_{\rm max}} \chi^{n}_{\rm f}\, \tau^{(j)}\,,
\end{align}
\end{subequations}
%
where we pull out from the sum all non-GR corrections, restricting
ourselves to the nonspinning ($j=0$) and linear-order in spin ($j=1$)
corrections to the GR QNMs.

\begin{table*}[t]
\begin{tabular}{c | c c c c c c c c c c}
\hline
\hline
Theory      & $\omega^{(0)}_{220}$ & $\tau^{(0)}_{220}$ & $\omega^{(1)}_{220}$ & $\tau^{(1)}_{220}$ & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & Ref.  \\
\hline
GR          & 0.3737               & 11.2407            &  0.1258              & 0.2522             & --  & --                          & --                        & --                          & --                        & \cite{Maselli:2019mjd} \\
\hline
EdGB        & --                   & --                 &  --                  & --                 & 4   & 0.0107                      & 0.0044                    & $-$0.2480                   & $-$1.101                  & \cite{Pani:2009wy,Blazquez-Salcedo:2016enn,Pierini:2021jxd} \\
dCS         & --                   & --                 &  --                  & --                 & 4   & 3.1964                      & 6.3619                    & 41.199                      & 794.66                    & \cite{Wagle:2021tam}  \\
cubic EFT   & --                   & --                 &  --                  & --                 & 4   & $-$0.5813                   & 2.6469                    & $-$3.8620                   & 265.12                    & \cite{Cano:2021myl} \\
quartic EFT & --                   & --                 &  --                  & --                 & 6   & $-$0.2114                   & $-$0.6070                 & $-$1.5263                   & 171.35                    & \cite{Cano:2021myl}  \\
\hline
\hline
\end{tabular}
\caption{Summary of theory-specific QNM calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(n)}_{220}$ and decay time $\delta \tau^{(n)}_{220}$, and the
references from which we used the results from.
%
We also included for comparison the GR coefficients, up to $j=1$, obtained in Ref.~\cite{Maselli:2019mjd}.
%
The remaining GR coefficients for $1< j \leqslant 4$, for which their non-GR counterpart cannot be determined as of yet for
the theories under consideration, can be found in Table~I of Ref.~\cite{Maselli:2019mjd}.
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

As we have seen in Sec.~\ref{sec:review_theories}, QNMs of slowly-rotating
BHs in modified gravity theories can belong to two families depending on how
they behave under a parity transformation: axial and polar. Which one do
we use to match with Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work with a chosen theory and perform a
translation between the metric perturbations $h_{\mu\nu}$ in the
Regge-Wheeler-Zerilli gauge~\cite{Regge:1957td,Zerilli:1970se} and connect it with the transverse-traceless gauge
used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Chapter~12),
%
In GR, both axial and polar QNMs are isospectral and hence which QNM we use to
model the ringdown makes no difference.
%
In non-GR theories, isospectrality is in general broken (see
in Ref.~\cite{Hui:2021cpm} for a counterexample).
%
Thus, how axial and polar gravitational QNMs appear in the GW signal has to be
answered on a theory-by-theory basis.
%
This is outside the scope of this paper and here we take the more pragmatic
approach of simply choosing the \emph{least damped gravitational mode between
the two parities.}
%
The justification is that this is the mode (if excited during the merger) which is
the most likely to appear in the signal.
We performed the mapping between theory-specific QNM calculations and the ParSpec framework
under the hypothesis above, for the theories listed in Sec.~\ref{sec:review_theories}.
%
We summarize our results in Table~\ref{tab:ref_theories_qnms} and
leave the details of our calculations to Appendix~\ref{app:map_details}.

@@

The \pSEOB{} model, as described in Sec.~\ref{sec:review_pSEOB}, is an
IMR model that infers the properties of the underlying GW signal,
including (independently) its ringdown properties, using the Bayesian formalism
above. Naturally, the most promising candidates for our analyses are high-mass
\emph{and} loud GW observations with a significant signal-to-noise ratio (SNR)
in the post-merger part of the signal.
%
The latest LVK GW catalog~\cite{LIGOScientific:2021djp} reported 90 observed
signals not all of which are relevant for our BH ringdown analysis. In fact, in
the accompanying paper~\cite{LIGOScientific:2021sio} on tests of GR, the \texttt{pSEOBNRv4HM}
~\cite{Brito:2018rfr,Ghosh:2021mrv} analysis~\footnote{See, in particular, Sec.VIII A.2
in Ref.~\cite{LIGOScientific:2021sio}}, which is most similar to the \pSEOB{} model
presented in this paper, identified two events which provided the strongest
bounds on the measurements of the dominant $(220)$ QNM:
GW150914~\cite{LIGOScientific:2016aoc} and
GW200129~\cite{LIGOScientific:2021djp}.
%
These two events, with a total (source-frame) mass of $65 \Mo$ and $63.4 \Mo$ respectively, are
extremely similar in their source properties. These are also two of the loudest
BBH signals observed to date with a total network SNR of 24 and 26.8,
respectively.
%
Moreover, and what is more relevant for our analysis, are their post-inspiral
(merger-ringdown) SNRs which are both $\sim `6$ (see the columns for
$\rho_{\text{post-insp}}$ in Table III of Ref.~\cite{LIGOScientific:2019fpa} and
Table IV of Ref.~\cite{LIGOScientific:2021sio}).
%
In this paper, we are going to focus on these two GW events as our probes of
the BH ringdown in modified theories of gravity.

@@

Let us also contrasts our results with those of Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}
which relied on the BBH inspiral to constrain $\ell_{\rm EdGB}$, as discussed in Sec.~\ref{sec:review_edgb}.
%
We see that EdGB gravity provides an example of a theory in which, with current GW events, the inspiral portion
of the signal can be more constraining than the ringdown portion of the signal.
%
Two reasons together can explain our negative results. First, as observed by
Ref.~\cite{Blazquez-Salcedo:2016enn}, the QNMs of EdGB BHs only differ slightly
from their Schwarzschild counterparts. Second, the larger mass $M_{\rm f}$ of the
remnant BH, suppresses scalar field's charge relatively to the initial binary
components.~\hscomm{This is a general `limitation' of ringdown test of higher-curvature theories in present detectors.
Point this out in the conclusions after all specific theories are discussed.}
%
As we will next, the situation is the opposite for dCS gravity.

@@

We presented an unified framework to combine the ParSpec formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LVK events GW150914 and GW200129.
%
We found, in particular that, within the interpretation of our non-GR theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm CS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm GB}$.
%
This dichotomy between these theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS which is \ab{poorly constrained with current observed GW
events}, while EdGB can be constrained, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
\hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

@@

@@

\begin{figure*}[t]
\includegraphics[width=0.9\textwidth]{figs/tmp_GW150914_intrinsic_params.pdf}
\caption{(Color online).~Corner plot showing that the inferred source binary parameters for GW150914,
using the same waveform model \pSEOB, but with (purple contours) and without setting the non-GR parameters different
from zero (blue contours), having as an illustrative example dCS gravity and $N_{\rm max}=0$.
%
\hs{Here, $\chi_{\rm eff}$ is the dimensionless effective-spin parameter, related to the individual spins $\chi_{i}$
and masses $m_{i}$ of each binary component as $\chi_{\rm eff} \equiv (m_1 \chi_1 + m_2 \chi_2) / (m_1 + m_2)$.}
%
All contours correspond to 90\% credible levels.
%
We see that the addition of the non-GR parameters does not introduce biases in the
inference of the source parameters.
%
We found the same qualitative behavior in the posteriors distributions of the
source binary parameters for the other modified gravity theories studied in the
main text.
%
The same conclusions apply for the other GW event studied here, GW200129.
}
\label{fig:corner_plot_all}
\end{figure*}

@@

We start by considering EdGB gravity, for which the QNMs of nonrotating and
slowly-rotating BHs where calculated in Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn,Blazquez-Salcedo:2017txk}
and~\cite{Pierini:2021jxd}, respectively.
%
In particular, Ref.~\cite{Blazquez-Salcedo:2016enn} found that the damping time
of the polar gravitational-led modes increases as the coupling constant
$\ell_{\rm EdGB}$ is increased.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm},
this is the gravitational-led branch of QNMs we select.

For the nonrotating QNMs we used the numerical data of Ref.~\cite{Blazquez-Salcedo:2016enn} and
generated a new quartic in $\ell_{\rm EdCS}$ fit to the data valid for small
values of the coupling constant [see, in particular, Eq.~(27) and Fig.~2 of~\cite{Blazquez-Salcedo:2016enn}].
%
We find,
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= M {\rm Re}(\sigma)_{\rm GR} \, \left[ 1 + 0.0107 (\ell_{\rm EdGB} / M)^4 \right],
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= M {\rm Im}(\sigma)_{\rm GR} \, \left[ 1 - 0.0044 (\ell_{\rm EdGB} / M)^4 \right].
\end{align}
\end{subequations}
%
The smallness of the numerical prefactors of $\ell_{\rm EdGB}^4$ are a consequence of the how weakly the QNMs
of BHs in EdGB gravity deviate from their GR counterparts, even at moderate values $(\ell_{\rm EdGB} / M)^4 \approx 0.3$.

The rotating corrections are included using the results of Ref.~\cite{Pierini:2021jxd}, in particular their Eqs.~(51) and~(52),
which for consistency, with our previous discussion, we truncated at quartic-order in $\ell_{\rm EdGB}$.
%
Taken together, all use as our fitting expressions for the QNM is EdGB gravity
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= \dots % {\rm Re}(\omega)_{\rm GR} \, \left[ 1 + 0.0107 (\ell_{\rm EdGB} / M)^4 \right],
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= \dots % {\rm Im}(\omega)_{\rm GR} \, \left[ 1 - 0.0044 (\ell_{\rm EdGB} / M)^4 \right].
\end{align}
\end{subequations}

@@

For dCS gravity, we follow Ref.~\cite{Wagle:2021tam} which numerically calculated the QNMs of slowly-rotating BHs in this theory
and found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm dCS}$ at fixed spin $\chi$ of the BH.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm},
this is the branch of QNMs we will select.

>>>>

A general procedure to describe deviations to the QNM frequencies $\omega^{\rm GR}_{\ell m n}$ and
damping times $\tau^{\rm GR}_{\ell m n}$ of BHs of GR is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{\ell m n} &= \omega_{\ell m n}^{\rm GR} \, (1 + \delta \omega_{\ell m n}), \\
\tau_{\ell m n }   &= \tau_{\ell m  n}^{\rm GR}   \, (1 + \delta \tau_{\ell m n }),
\end{align}
\end{subequations}
%
where $\delta\omega_{\ell m n}$ and $\delta\tau_{\ell m  n}$ are the fractional deformation parameters,
$\ell$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization\footnote{See also
Refs.~\cite{Glampedakis:2017dvb,Glampedakis:2019dqh,Silva:2019scu} and
Refs.~\cite{Cardoso:2019mqo,McManus:2019ulj,Kimura:2020mrh} for alternative parametrizations.}
was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Brito:2018rfr,Carullo:2018sfu,Ghosh:2021mrv}.

@@

The ParSpec framework was introduced in Ref.~\cite{Maselli:2019mjd} and can be
used to our purpose. It is an observable-based bivariate expansion of
Eq.~\eqref{eq:general_deviation}, given by
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{\ell m n} &= \frac{1}{M_{\rm f}} \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \omega^{(j)}_{\ell m n} \, \left( 1 + \gamma\, \delta \omega^{(j)}_{\ell m n} \right), \\
\tau_{\ell m n}   &= M_{\rm f}     \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \tau^{(j)}_{\ell m n}   \, \left( 1 + \gamma\, \delta \tau^{(j)}_{\ell m n} \right),
\end{align}
\end{subequations}
%
where $M_{\rm f}$ and $\chi_{\rm f}$ are the detector-frame final mass and spin, respectively;
the quantities $\omega_{\ell m n}^{(j)}$ and $\tau_{\ell m n}^{(j)}$ are dimensionless coefficients of the
expansion in spin for the QNMs of BHs in GR, while $\delta\omega_{\ell m n}^{(j)}$ and $\delta\tau_{\ell m n}^{(j)}$ are
source-independent dimensionless coefficients that characterize the corrections to the GR QNM at
each spin-order,
%
and $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence due to a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, which reads
%
\begin{equation}
\gamma
%
= \left(\frac{\ell_{\rm th}}{M^{\rm s}_{\rm f}}\right)^{p}
%
= \left[
\frac{\ell_{\rm th} c^2 (1 + z)}{G M_{\rm f}}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on the length-scale parameter $\ell_{\rm th}$ of the specific gravity theory
(non-GR modifications become important at distances $\lesssim \ell_{\rm th}$), and
the exponent $p$ is related to how the non-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\gamma$ dimensionless by the
length-scale associated with the remnant BH (i.e., its source-frame mass
$M^{\rm s}_{\rm f}$\footnote{Throughout this paper, we maintain the convention of
using the upperscript ``s" to denote source-frame quantities and plain baseline
symbols for detector-frame measurements.}), which we can also write in terms of the detector-frame mass $M_{\rm f}$
through the redshift $z$~\cite{Krolak:1987ofj}. Also, dividing by the factor $G/c^2$ allows us to
express $\ell_{\rm th}$ in physically intuitive metric units.

@@

We now describe the waveform model used in our paper to infer properties of a
BBH ringdown.
%
As in Refs.~\cite{Brito:2018rfr,Ghosh:2021mrv}, we use an IMR BBH waveform
model where the complex-valued frequencies describing the remnant object
are left additionally free and estimated directly from the data.

@@

Let us now establish the connection between the theory-independent framework of the
$\pSEOB$ waveform model and the theory-specific QNM calculations. In this paper we restrict ourselves
to the leading and next-to-leading order terms in the ParSpec expansion, as well as to the fundamental QNM
$(\ell, m, n) = (2,\, 2,\, 0)$. For this reason, for simplicity, we omit the subscripts hereafter
and rewrite $\omega_{\ell m n}$ and $\tau_{\ell m n}$, given by Eqs.~\eqref{eq:kerr_expansion} as,
%
\begin{subequations}
\label{eq:qnm_nongr_iso}
\begin{align}
    M_{\rm f}\, \omega &= \gamma \left [ \delta\omega^{(0)} \omega^{(0)} + \chi_{\rm f}\, \delta\omega^{(1)} \omega^{(1)} \right ]
    + \sum_{j=0}^{N_{\rm max}} \chi^{j}_{\rm f}\, \omega^{(j)} \,,
\\
    \frac{\tau}{M_{\rm f}}   &= \gamma \left [ \delta\tau^{(0)} \tau^{(0)} + \chi_{\rm f}\, \delta\tau^{(1)} \tau^{(1)} \right ]
    + \sum_{j=0}^{N_{\rm max}} \chi^{n}_{\rm f}\, \tau^{(j)}\,,
\end{align}
\end{subequations}
%
where we pull out from the sum all non-GR corrections, restricting
ourselves to the nonspinning ($j=0$) and linear-order in spin ($j=1$)
corrections to the GR QNMs.

\begin{table*}[t]
\begin{tabular}{c | c c c c c c c c c c}
\hline
\hline
Theory      & $\omega^{(0)}_{220}$ & $\tau^{(0)}_{220}$ & $\omega^{(1)}_{220}$ & $\tau^{(1)}_{220}$ & $p$ & $\delta \omega^{(0)}_{220}$ & $\delta \tau^{(0)}_{220}$ & $\delta \omega^{(1)}_{220}$ & $\delta \tau^{(1)}_{220}$ & Ref.  \\
\hline
GR          & 0.3737               & 11.2407            &  0.1258              & 0.2522             & --  & --                          & --                        & --                          & --                        & \cite{Maselli:2019mjd} \\
\hline
EdGB        & --                   & --                 &  --                  & --                 & 4   & 0.0107                      & 0.0044                    & $-$0.2480                   & $-$1.101                  & \cite{Blazquez-Salcedo:2016enn,Pierini:2021jxd} \\
dCS         & --                   & --                 &  --                  & --                 & 4   & 3.1964                      & 6.3619                    & 41.199                      & 794.66                    & \cite{Wagle:2021tam}  \\
cubic EFT   & --                   & --                 &  --                  & --                 & 4   & $-$0.5813                   & 2.6469                    & $-$3.8620                   & 265.12                    & \cite{Cano:2021myl} \\
quartic EFT & --                   & --                 &  --                  & --                 & 6   & $-$0.2114                   & $-$0.6070                 & $-$1.5263                   & 171.35                    & \cite{Cano:2021myl}  \\
\hline
\hline
\end{tabular}
\caption{Summary of theory-specific QNM calculations.
%
We summarize each theory we have considered together with: the exponent $p$ at
which their QNM-modification enters, the corresponding modifications to the
oscillation frequency $\delta \omega^{(n)}_{220}$ and decay time $\delta \tau^{(n)}_{220}$, and the
references from which we used the results from.
%
We also included for comparison the GR coefficients, up to $j=1$, obtained in Ref.~\cite{Maselli:2019mjd}.
%
The remaining GR coefficients for $1< j \leqslant 4$, for which their non-GR counterpart cannot be determined as of yet for
the theories under consideration, can be found in Table~I of Ref.~\cite{Maselli:2019mjd}.
}
\label{tab:ref_theories_qnms}
\end{table*}

@@

As we have seen in Sec.~\ref{sec:review_theories}, QNMs of slowly-rotating
BHs in modified gravity theories can belong to two families depending on how
they behave under a parity transformation: axial and polar. Which one do
we use to match with Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work with a chosen theory and perform a
translation between the metric perturbations $h_{\mu\nu}$ in the
Regge-Wheeler-Zerilli gauge~\cite{Regge:1957td,Zerilli:1970se} and connect it with the transverse-traceless gauge
used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Chapter~12),
%
In GR, both axial and polar QNMs are isospectral and hence which QNM we use to
model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see
in Ref.~\cite{Hui:2021cpm} for a counterexample).
%
Thus, how axial and polar gravitational QNMs appear in the GW signal has to be
answered on a theory-by-theory basis.
%
This is outside the scope of this paper and here we take the more pragmatic
approach of simply choosing the \emph{least damped gravitational mode between
the two parities.}
%
The justification is that this is the mode (if excited during the merger) which is
the most likely to appear in the signal.
We performed the mapping between theory-specific QNM calculations and the ParSpec framework
under the hypothesis above, for the theories listed in Sec.~\ref{sec:review_theories}.
%
We summarize our results in Table~\ref{tab:ref_theories_qnms} and
leave the details of our calculations to Appendix~\ref{app:map_details}.

@@

The \pSEOB{} model, as described in Sec.~\ref{sec:review_pSEOB}, is an
IMR model that infers the properties of the underlying GW signal,
including (independently) its ringdown properties, using the Bayesian formalism
above. Naturally, the most promising candidates for our analyses are high-mass
\emph{and} loud GW observations with a significant signal-to-noise ratio (SNR)
in the post-merger part of the signal.
%
The latest LVK GW catalog~\cite{LIGOScientific:2021djp} reported 90 observed
signals not all of which are relevant for our BH ringdown analysis. In fact, in
the accompanying paper~\cite{LIGOScientific:2021sio} on tests of GR, the \texttt{pSEOBNRv4HM}
~\cite{Brito:2018rfr,Ghosh:2021mrv} analysis~\footnote{See, in particular, Sec.VIII A.2
in Ref.~\cite{LIGOScientific:2021sio}}, which is most similar to the \pSEOB{} model
presented in this paper, identified two events which provided the strongest
bounds on the measurements of the dominant $(220)$ QNM:
GW150914~\cite{LIGOScientific:2016aoc} and
GW200129~\cite{LIGOScientific:2021djp}.
%
These two events, with a total (source-frame) mass of $65 \Mo$ and $63.4 \Mo$ respectively, are
extremely similar in their source properties. These are also two of the loudest
BBH signals observed to date with a total network SNR of 24 and 26.8,
respectively.
%
Moreover, and what is more relevant for our analysis, are their post-inspiral
(merger-ringdown) SNRs which are both $\approx 6$ (see the columns for
$\rho_{\text{post-insp}}$ in Table III of Ref.~\cite{LIGOScientific:2019fpa} and
Table IV of Ref.~\cite{LIGOScientific:2021sio}).
%
In this paper, we are going to focus on these two GW events as our probes of
the BH ringdown in modified theories of gravity.

@@

Let us also contrasts our results with those of Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}
which relied on the BBH inspiral to constrain $\ell_{\rm EdGB}$, as discussed in Sec.~\ref{sec:review_edgb}.
%
We see that EdGB gravity provides an example of a theory in which, with current GW events, the inspiral portion
of the signal can be more constraining than the ringdown portion of the signal.
%
Two reasons together can explain our negative results. First, as observed by
Ref.~\cite{Blazquez-Salcedo:2016enn}, the QNMs of EdGB BHs only differ slightly
from their Schwarzschild counterparts. Second, the larger mass $M_{\rm f}$ of the
remnant BH, suppresses scalar field's charge relatively to the initial binary
components.
%
As we will next, the situation is the opposite for dCS gravity.

@@

We presented an unified framework that combines the ParSpec formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LVK events GW150914 and GW200129.
%
We found, in particular that, within the interpretation of our non-GR theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm EdGB}$.
%
This dichotomy between these theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS which is \ab{poorly constrained with current observed GW
events}, while EdGB can be constrained, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
\hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

@@

~\hscomm{This is a general `limitation' of ringdown test of higher-curvature theories in present detectors.
Point this out in the conclusions after all specific theories are discussed.}

@@

\begin{figure*}[t]
\includegraphics[width=0.9\textwidth]{figs/tmp_GW150914_intrinsic_params.pdf}
\caption{(Color online).~Corner plot showing that the inferred source binary parameters for GW150914,
using the same waveform model \pSEOB, but with (purple contours) and without setting the non-GR parameters different
from zero (blue contours), having as an illustrative example dCS gravity and $N_{\rm max}=0$.
%
Here, $\chi_{\rm eff}$ is the dimensionless effective-spin parameter, related to the individual spins $\chi_{i}$
and masses $m_{i}$ of each binary component as $\chi_{\rm eff} \equiv (m_1 \chi_1 + m_2 \chi_2) / (m_1 + m_2)$.
%
All contours correspond to 90\% credible levels.
%
We see that the addition of the non-GR parameters does not introduce biases in the
inference of the source parameters.
%
We found the same qualitative behavior in the posteriors distributions of the
source binary parameters for the other modified gravity theories studied in the
main text.
%
The same conclusions apply for the other GW event studied here, GW200129.
}
\label{fig:corner_plot_all}
\end{figure*}

@@

We start by considering EdGB gravity and, in particular, focus on
Refs.~\cite{Blazquez-Salcedo:2016enn,Pierini:2021jxd} to determine the
ParSpec coefficients for this theory.
% and~\cite{Pierini:2021jxd} for the QNMs of for which the QNMs of nonrotating and
% slowly-rotating BHs where calculated in Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn,Blazquez-Salcedo:2017txk}
% and~\cite{Pierini:2021jxd}, respectively.
%
In particular, Ref.~\cite{Blazquez-Salcedo:2016enn} found that the damping time
of the polar gravitational-led modes increases as the coupling constant
$\ell_{\rm EdGB}$ is increased. The leading-order spin corrections to the
branch of QNMs was calculated in~\cite{Pierini:2021jxd}.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm},
this is the gravitational-led branch of QNMs we select.

For the nonrotating QNMs we used the numerical data of Ref.~\cite{Blazquez-Salcedo:2016enn} and
generated a new quartic fit in $\gamma_{\rm EdGB} = \ell_{\rm EdCS} / M$ using numerical QNM data valid for small
values of the coupling constant [see, in particular, Eq.~(27) and Fig.~2 of Ref.~\cite{Blazquez-Salcedo:2016enn}].
%
We find,
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= M {\rm Re}(\sigma)_{\rm GR} \, \left( 1 + 0.0107 \, \gamma_{\rm EdGB}^4 \right),
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= M {\rm Im}(\sigma)_{\rm GR} \, \left( 1 - 0.0044 \, \gamma_{\rm EdGB}^4 \right).
\end{align}
\end{subequations}
%
The small values of the numerical prefactors of $\gamma_{\rm EdGB}^4$ are a consequence of the how weakly the QNMs
of BHs in EdGB gravity deviate from their GR counterparts, even at moderate values $(\ell_{\rm EdGB} / M)^4 \approx 0.3$.

The rotating corrections are included using the results of Ref.~\cite{Pierini:2021jxd}, in particular their Eqs.~(51) and~(52),
which for consistency with our previous discussion, we truncated at quartic-order in $\gamma_{\rm EdGB}$.
%
Taken together, use as our fitting expressions for the QNM is EdGB gravity
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= {\rm Re}(\omega)_{\rm GR} \, \left[ 1 + 0.0107 (\ell_{\rm EdGB} / M)^4 \right],
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= \dots % {\rm Im}(\omega)_{\rm GR} \, \left[ 1 - 0.0044 (\ell_{\rm EdGB} / M)^4 \right].
\end{align}
\end{subequations}

@@

For dCS gravity, we follow Ref.~\cite{Wagle:2021tam} which numerically calculated the QNMs of slowly-rotating BHs in this theory
and found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm dCS}$ at fixed spin $\chi$ of the BH.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm},
this is the branch of QNMs we chose to work with.
